plyranges: A grammar of genomic data transformation

### 2018-09-20 paper.Rmd
spelling and typo checks

---
title: "plyranges: A grammar of genomic data transformation" 
author:
  - name: Stuart Lee \footnote{\label{monash}Department of Econometrics and Business Statistics, Monash University, Clayton, Australia}\footnote{Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia}
    email: stuart.lee1@monash.edu
  - name: Dianne Cook $^{\ref{monash}}$
    email: di.cook@monash.edu 
  - name: Michael Lawrence \footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author} 
    email: lawrence.michael@gene.com
abstract: "The Bioconductor project provides many interoperable data abstractions for analysing high-throughput genomics experiments; however implementing a typical genomic workflow with Bioconductor requires learning these abstractions and understanding them at an integrative level. This places a large cognitive burden on the user, especially for non-programmers. To reduce this burden we have created a grammar of genomic data transformation that operates on a single, central Bioconductor data structure, GRanges, which naturally represents genomic intervals and their associated measurements. The grammar defines verbs for performing actions on and between genomic interval data through a simplified, coherent interface to existing Bioconductor infrastructure, resulting in fluent analysis workflows. We have implemented this grammar as an R/Bioconductor package called \\texttt{plyranges}."
keywords: "Bioconductor, Grammar, Genomes, Data Analysis" 
bibliography: references.bib 
output: 
  pdf_document:
    template: svm-latex-ms.tex 
    keep_tex: true 
    fig_caption: true
    latex_engine: pdflatex 
    citation_package: biblatex 
fontfamily: mathpazo
---

@@

The Bioconductor/R packages `IRanges` and `GenomicRanges` 
[@r-core; @Lawrence2013-wg; @Huber2015-ei] define a DSL for 
analysing genomics data with R, an interactive data analysis environment that 
encourages reproducibility and provides high-level abstractions for manipulating,
modelling and plotting data, through state of the art methods in
statistical computing. The packages define object-oriented (OO)
abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. Other genomic DSLs 
that are embedded in programming languages include pybedtools and valr 
[@Dale2011-js; @Kent2017], however these packages lack the interoperability
provided by the aforementioned Bioconductor packages and are not easily 
extended.

@@

\begin{table}[!htbp]
\centering
\begin{tabular}{|l|l|p{6cm}|}
  \hline
  & Verb &  Description \\ 
  \hline
   & \textbf{\emph{summarise()}} & aggregate over column(s) \\ 
   Aggregate & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighbouring ranges \\
   \hline
   &  \textbf{\emph{mutate()}} & modifies any column \\
   & \textbf{\emph{select()}} & select columns \\
  Modify (Unary) & \textbf{\emph{arrange()}} & sort by columns \\
   & \emph{stretch()} & extend range by fixed amount \\
   &  \emph{shift\_(direction)} & shift coordinates \\
   & \emph{flank\_(direction)} & generate flanking regions \\
   & \emph{\%intersection\% } & row-wise intersection \\
   & \emph{\%union\%} & row-wise union \\
   & \emph{compute\_coverage} & coverage over all ranges \\
  Modify (Binary) &  \emph{\%setdiff\%} & row-wise set difference \\
   & \emph{between()} & row-wise gap range \\
   & \emph{span()} & row-wise spanning range \\
   \hline
    & \emph{join\_overlap\_*()} & merge by overlapping ranges \\
    & \emph{join\_nearest} & merge by nearest neighbour ranges \\
    & \emph{join\_follow} & merge by following ranges \\
    Merge & \emph{join\_precedes} & merge by preceding ranges \\
    & \emph{union\_ranges} & range-wise union \\
    & \emph{intersect\_ranges} & range-wise intersect \\
    & \emph{setdiff\_ranges} & range-wise set difference \\
    & \emph{complement\_ranges} & range-wise union \\
  \hline
   & \emph{anchor\_direction()} & fix coordinates at direction \\
  Operate & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
   & \emph{group\_by\_overlaps()} & partition by overlaps \\
   \hline
   & \textbf{\emph{filter()}} & subset rows \\
  Restrict & \emph{filter\_by\_overlaps()} & subset by overlap \\
    & \emph{filter\_by\_non\_overlaps()} & subset by no overlap \\
   \hline
\end{tabular}
\caption{Overview of the \texttt{plyranges} grammar. The core verbs are
briefly described and categorised into one of the following higher level 
catgeories: aggregate, modify, merge, operate, or restrict. A verb is given bold text if
its origin is from the \texttt{dplyr} grammar.}\label{tab:grammar}
\end{table}

@@

\begin{figure}
\includegraphics[width=\textwidth]{diagrams/olap-fig}
\caption{Illustration of the three overlap join operators. Each join takes two 
GRanges objects, \textit{x} and \textit{y} as input. A 'Hits' object for the 
join is computed which consists of two components. The first component contains 
the indices of the ranges in \textit{x} that have been overlapped 
(the rectangles of \textit{x} that cross the orange lines). The second 
component consists of the indices of the ranges in \textit{y} that overlap 
the ranges in \textit{x}. In this case a range in \textit{y} overlaps the 
ranges in \textit{x} three times, so the index is repeated three times. The 
resulting 'Hits' object is used to modify \textit{x} by where it was 'hit' by
\textit{y} and merge all metadata columns from \textit{x} and \textit{y} 
based on the indices contained in the 'Hits' object. This procedure is applied 
generally in the \texttt{plyranges} DSL for both overlap and nearest neighbour
operations. The join semantics alter what is returned: \textbf{A}: for an 
\textbf{inner} join the \textit{x} ranges that are overlapped by \textit{y} 
are returned. The returned ranges also include the metadata from the \textit{y}
range that overlapped the three \textit{x} ranges. \textbf{B} An 
\textbf{intersect} join is identical to an inner join except that the 
intersection is taken between the overlapped \textit{x} ranges and the 
\textit{y} ranges. \textbf{C} For the left join \textbf{left} join all 
\textit{x} ranges are returned regardless of whether they are overlapped by 
\textit{y}. In this case the resulting third range  of the join would have 
missing values on metadata columns that came from \textit{y}.}
\label{fig:olaps-fig}
\end{figure}

@@

Here we provide illustrative examples of using the `plyranges` DSL to show
how our grammar could be integrated into genomic data workflows. As we
construct the workflows we show the data output intermittantly to assist
the reader in understanding the pipeline steps. The workflows
highlight how interoperability with existing Bioconductor infrastructure, 
enables easy access to public datasets and methods for analysis and 
visualization.

@@

The implementation of `plyranges` is built on top of 
Bioconductor infrastructure, meaning most functions are constructed
by composing generic functions from core Bioconductor packages. As a result, any 
Bioconductor packages that uses data structures that inherit from GRanges will
be able to use `plyranges`  for free. Another consequence of building on top of
Bioconductor generics is that the speed and memory usage of `plyranges` 
functions are similar to the highly optimised methods implemented in Bioconductor
for GRanges objects.

@@

We aim to continue developing the `plyranges` package and to extend it
for use with more complex data structures, such as the
SummarizedExperiment class, the core Bioconductor data structure for
representing experimental results (e.g., counts) from multiple sample
experiments in conjunction with feature and sample metadata. Although,
the SummarizedExperiment is not strictly tidy, it does consist of 
three tidy data structures that are related by feature and sample identifiers. 
Thererfore, the grammar and design of the `plryanges` DSL is naturally 
extensible to the SummarizedExperiment.

>>>>

---
title: "plyranges: A grammar of genomic data transformation" 
author:
  - name: Stuart Lee \footnote{\label{monash}Department of Econometrics and Business Statistics, Monash University, Clayton, Australia}\footnote{Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia}
    email: stuart.lee1@monash.edu
  - name: Dianne Cook $^{\ref{monash}}$
    email: di.cook@monash.edu 
  - name: Michael Lawrence \footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author} 
    email: lawrence.michael@gene.com
abstract: "The Bioconductor project provides many interoperable data abstractions for analyzing high-throughput genomics experiments; however implementing a typical genomic workflow with Bioconductor requires learning these abstractions and understanding them at an integrative level. This places a large cognitive burden on the user, especially for non-programmers. To reduce this burden we have created a grammar of genomic data transformation that operates on a single, central Bioconductor data structure, GRanges, which naturally represents genomic intervals and their associated measurements. The grammar defines verbs for performing actions on and between genomic interval data through a simplified, coherent interface to existing Bioconductor infrastructure, resulting in fluent analysis workflows. We have implemented this grammar as an R/Bioconductor package called \\texttt{plyranges}."
keywords: "Bioconductor, Grammar, Genomes, Data Analysis" 
bibliography: references.bib 
output: 
  pdf_document:
    template: svm-latex-ms.tex 
    keep_tex: true 
    fig_caption: true
    latex_engine: pdflatex 
    citation_package: biblatex 
fontfamily: mathpazo
---

@@

The Bioconductor/R packages `IRanges` and `GenomicRanges` 
[@r-core; @Lawrence2013-wg; @Huber2015-ei] define a DSL for 
analyzing genomics data with R, an interactive data analysis environment that 
encourages reproducibility and provides high-level abstractions for manipulating,
modelling and plotting data, through state of the art methods in
statistical computing. The packages define object-oriented (OO)
abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. Other genomic DSLs 
that are embedded in programming languages include pybedtools and valr 
[@Dale2011-js; @Kent2017], however these packages lack the interoperability
provided by the aforementioned Bioconductor packages and are not easily 
extended.

@@

\begin{table}[!htbp]
\centering
\begin{tabular}{|l|l|p{6cm}|}
  \hline
  & Verb &  Description \\ 
  \hline
   & \textbf{\emph{summarize()}} & aggregate over column(s) \\ 
   Aggregate & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighboring ranges \\
   \hline
   &  \textbf{\emph{mutate()}} & modifies any column \\
   & \textbf{\emph{select()}} & select columns \\
  Modify (Unary) & \textbf{\emph{arrange()}} & sort by columns \\
   & \emph{stretch()} & extend range by fixed amount \\
   &  \emph{shift\_(direction)} & shift coordinates \\
   & \emph{flank\_(direction)} & generate flanking regions \\
   & \emph{\%intersection\% } & row-wise intersection \\
   & \emph{\%union\%} & row-wise union \\
   & \emph{compute\_coverage} & coverage over all ranges \\
  Modify (Binary) &  \emph{\%setdiff\%} & row-wise set difference \\
   & \emph{between()} & row-wise gap range \\
   & \emph{span()} & row-wise spanning range \\
   \hline
    & \emph{join\_overlap\_*()} & merge by overlapping ranges \\
    & \emph{join\_nearest} & merge by nearest neighbor ranges \\
    & \emph{join\_follow} & merge by following ranges \\
    Merge & \emph{join\_precedes} & merge by preceding ranges \\
    & \emph{union\_ranges} & range-wise union \\
    & \emph{intersect\_ranges} & range-wise intersect \\
    & \emph{setdiff\_ranges} & range-wise set difference \\
    & \emph{complement\_ranges} & range-wise union \\
  \hline
   & \emph{anchor\_direction()} & fix coordinates at direction \\
  Operate & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
   & \emph{group\_by\_overlaps()} & partition by overlaps \\
   \hline
   & \textbf{\emph{filter()}} & subset rows \\
  Restrict & \emph{filter\_by\_overlaps()} & subset by overlap \\
    & \emph{filter\_by\_non\_overlaps()} & subset by no overlap \\
   \hline
\end{tabular}
\caption{Overview of the \texttt{plyranges} grammar. The core verbs are
briefly described and categorized into one of the following higher level 
catgeories: aggregate, modify, merge, operate, or restrict. A verb is given bold text if
its origin is from the \texttt{dplyr} grammar.}\label{tab:grammar}
\end{table}

@@

\begin{figure}
\includegraphics[width=\textwidth]{diagrams/olap-fig}
\caption{Illustration of the three overlap join operators. Each join takes two 
GRanges objects, \textit{x} and \textit{y} as input. A 'Hits' object for the 
join is computed which consists of two components. The first component contains 
the indices of the ranges in \textit{x} that have been overlapped 
(the rectangles of \textit{x} that cross the orange lines). The second 
component consists of the indices of the ranges in \textit{y} that overlap 
the ranges in \textit{x}. In this case a range in \textit{y} overlaps the 
ranges in \textit{x} three times, so the index is repeated three times. The 
resulting 'Hits' object is used to modify \textit{x} by where it was 'hit' by
\textit{y} and merge all metadata columns from \textit{x} and \textit{y} 
based on the indices contained in the 'Hits' object. This procedure is applied 
generally in the \texttt{plyranges} DSL for both overlap and nearest neighbor
operations. The join semantics alter what is returned: \textbf{A}: for an 
\textbf{inner} join the \textit{x} ranges that are overlapped by \textit{y} 
are returned. The returned ranges also include the metadata from the \textit{y}
range that overlapped the three \textit{x} ranges. \textbf{B} An 
\textbf{intersect} join is identical to an inner join except that the 
intersection is taken between the overlapped \textit{x} ranges and the 
\textit{y} ranges. \textbf{C} For the \textbf{left} join all 
\textit{x} ranges are returned regardless of whether they are overlapped by 
\textit{y}. In this case the third range (rectangle with the asterisk next to it)
of the join would have missing values on metadata columns that came from \textit{y}.}
\label{fig:olaps-fig}
\end{figure}

@@

Here we provide illustrative examples of using the `plyranges` DSL to show
how our grammar could be integrated into genomic data workflows. As we
construct the workflows we show the data output intermittently to assist
the reader in understanding the pipeline steps. The workflows
highlight how interoperability with existing Bioconductor infrastructure, 
enables easy access to public datasets and methods for analysis and 
visualization.

@@

The implementation of `plyranges` is built on top of 
Bioconductor infrastructure, meaning most functions are constructed
by composing generic functions from core Bioconductor packages. As a result, any 
Bioconductor packages that uses data structures that inherit from GRanges will
be able to use `plyranges`  for free. Another consequence of building on top of
Bioconductor generics is that the speed and memory usage of `plyranges` 
functions are similar to the highly optimized methods implemented in Bioconductor
for GRanges objects.

@@

We aim to continue developing the `plyranges` package and to extend it
for use with more complex data structures, such as the
SummarizedExperiment class, the core Bioconductor data structure for
representing experimental results (e.g., counts) from multiple sample
experiments in conjunction with feature and sample metadata. Although,
the SummarizedExperiment is not strictly tidy, it does consist of 
three tidy data structures that are related by feature and sample identifiers. 
Therefore, the grammar and design of the `plryanges` DSL is naturally 
extensible to the SummarizedExperiment.
